---
title: "metaheuR: Optimization problems"
author: "Borja Calvo, Josu Ceberio, Usue Mori"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: sandstone
    highlight: pygments
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{First steps}
  %\VignetteEngine{knitr::docco_linear}
---

When working with an optimization problem there are two aspects that have to be defined: The solution representation and the evaluation function. Depending on the problem and the codification used, there could be solutions that are not feasible.

When we create an optimization problem in `metaheuR` we have to define, at least, a function to evaluate solutions. This function must have a single parameter `solution`, the solution to be evaluated, and must return a numeric value. The implementation of this function will depend on how the solutions are codified, so we need to be very clear about the kind of object that the function requires. 

As an example, the function `tspProblem` creates the evaluation function used to solve instances of the _Travelling Salesperson Problem_. In this problem, we have `n` cities and a matrix containing the cost of traveling between any pair of cities. The goal of this problem is to find the minimum cost circuit that passes through every single city once and only once. 

In order to create an instance of the problem, we need a cost matrix. For example, we can create a random instance of the problem using a random matrix.

```{r,message=FALSE}
library("metaheuR")
n <- 50
rnd.matrix <- matrix(runif(n^2), ncol=n)
rnd.tsp <- tspProblem(cmatrix=rnd.matrix)
names(rnd.tsp)
```

As can be seen in the code, the `tspProblem` function returns a list with a single function, named `evaluate`. This function evaluates any solution taking into account the cost matrix introduced in the `tspProblem` function. In this implementation of the problem, the solutions are represented by permutations. As it is quite a common representation, the package includes a class, `Permutation`, to handle this type of solution. We can now create a random solution using the `randomPermutation` function and evaluate it using the function we have just built.

```{r}
rnd.solution <- randomPermutation(length = n)
rnd.tsp$evaluate(rnd.solution)
```

The package also includes a function to load cost matrices from the well known [TSPlib repository](http://www.iwr.uni-heidelberg.de/groups/comopt/software/TSPLIB95/). This function can be used to load XML files and it accepts both local and remote files, either zip compressed or uncompressed.


```{r}

cost.matrix <- tsplibParser(file="http://www.iwr.uni-heidelberg.de/groups/comopt/software/TSPLIB95/XML-TSPLIB/instances/brazil58.xml.zip")
tsp.brazil <- tspProblem(cost.matrix)
```

Now that we have the evaluation function we can evaluate the identity permutation as a solution to the problem

```{r}
n <- dim(cost.matrix)[1]
csol <- identityPermutation(n)
tsp.brazil$evaluate(csol)
```

## Problems with restrictions

In some optimization problems there are solutions that may not be feasible. In such cases, the problem generating function must provide functions to handle them. In particular, these functions should provide two additional functions, `valid` and `correct`, with only one input parameter, `solution`. The first function should output `TRUE` if the introduced solution is valid and `FALSE` otherwise. The second function should modify non-valid solutions in some way in order to make them valid.

An example of this kind of problem is the Knapsack Problem. We have a knapsack and a list of items. Each item has a _weight_ and a _value_, and the knapsack has a maximum capacity. The goal is to fill the knapsack with items, maximizing the total value of them but without surpassing the limit of the knapsack. We can create a random knapsack problem generating two random vectors which correspond to the weights and values and a random value which defines the limit of the knapsack. To make it more realistic, the values will be correlated with the weights and the limit will be determined by summing the weight of some random items.

```{r}
n <- 10
weight <- runif(n)
value <- weight + 0.2 * runif(n)
capacity <- sum(sample(weight, size=0.25 * n, replace=FALSE))
ksp <- knapsackProblem(weight=weight, value=value, limit=capacity)
names(ksp)
```

The function `knapsackProblem` returns functions that can be used to check and correct any solution. We can test them with a solution that is obviously non-valid, the one that includes all the items. In this problem the solutions are represented as vectors of logic values, so the solution that includes all the items in the knapsack is a vector of size `r n` where all the values are `TRUE`.

```{r}
all.in.solution <- rep(TRUE, n)
ksp$valid(all.in.solution)
corrected.solution <- ksp$correct(all.in.solution)
ksp$valid(corrected.solution)
```

In this case, the non-valid solutions are corrected by consecutively removing the items that have the highest weight/value ratio, until the limit is respected.

In some other problems, unfeasible solutions are handled by adding a penalization term in the evaluation function. The package also incorporates a problem implementation that includes this posibility, the `misProblem` function.

This function creates the functions needed to solve Maximum Independece Set (MIS) problems. In these problems, given a graph, the goal is finding the biggest subset of nodes such that there is no link between any two nodes in the subset. Such a subset is called _independent set_, and the solution to the problem is the maximum one, that is, the biggest _independet set_. 

The `misProblem` function has two parameters, `graph` and `penalization`. The first parameter is the graph representing the instance of the problem. The second one is used to penalize non valid solutions, i.e., subsets of nodes that are not independent. The penalization term is computed as the product of the parameter introduced (`penalization`) and the number of nodes that violate the independence condition.

Let us see the effect of the penalization in an example based on a random graph.

```{r}
set.seed(1)
library(igraph)
n <- 100
rnd.graph <- random.graph.game (n=n, p.or.m=0.05)
misp <- misProblem(graph=rnd.graph, penalization=0)
misp.penalized <- misProblem(graph=rnd.graph, penalization=0.5)
```

We have created two versions of the problem, with and without penalization. Now we create a random subset of the nodes and check the evaluation assigned by each of the problems.

```{r}
rnd.solution <- runif(n) > 0.4
misp$valid(rnd.solution)
misp$evaluate(rnd.solution)
misp.penalized$evaluate(rnd.solution)
```

Note that the evaluations are negative. This is because the objective in MIS problems is maximizing the size of the independent set. However, in the package, all the algorithms are programmed to minimize the objective function, so in such situations we replace the original objective function $f$ (in this case, the size of the set) with a new function $g=-f$. We can see this by typing:

```{r}
misp$evaluate
```

Therefore, the smallest solution the better, that is, penalization terms will increase the value of the objective function. 

If we correct the solution, then it will become a feasible solution, it will not have any penalization and, thus, the evaluation will be the same with both versions of the problem.

```{r}
valid.solution <- misp$correct(rnd.solution)
misp$evaluate(valid.solution)
misp.penalized$evaluate(valid.solution)
```

## Additional functions

In some problem implementations we can find additional functions that may help in visualizing solutions. In the particular case of graph problems, the problem generating function returns a function called `plot`, that can be used to plot any solution in the actual graph.

```{r}
misp$plot(rnd.solution, node.size=5)
misp$plot(valid.solution)
```


## List of functions in the package

The complete list of problems implemented in the current vesion of the packages is the following

Problem | Function
--------|----------
Traveling Salesperson Problem | `tspProblem`
Knapsack Problem | `knpasackProblem`
Graph Colorig Problem | `graphColoringProblem`
Maximum Independent Set Problem | `misProblem`


## Implementing new problems

Anyone can implement new problems to be solved with the algoritms included in the package. In order to keep the coherence between implementations, it is advisable to follow the design pattern used to implement all the problems contained in the package. 

The design consists in a function that returns a list of functions. The benefit of this desing is that the function will encapsulate in its evironment all of its arguments and, thus, the functions returned in the list will have access to them.

Any problem has to provide, at least, an evaluation function (named `evaluation`) that will take a single argument (a solution) and will return a numeric value (its evaluation). Depending on the problem, we may need to return also a function to check the validity of a solution and a function to correct non valid solutions. The former will be named `valid`, will take as its only argument a solution and will return `TRUE` if the solution is correct or `FALSE` if it is incorrect. The later (named `correct`) will take as input a solution and will return the corrected solution (or the same solution, if it is already correct). This is the template to implement new problems:

```{r }
myProblem <- function(arg1, arg2, arg3) {
  
  evaluate <- function(solution) {
    return (-1)
  }
  
  valid <- function(solution) {
    return(TRUE)
  }
  
  correct <- function(solution) {
    return(solution)
  }
  
  return(list(evaluate=evaluate, valid=valid, correct=correct))
}
```

Note that the above is the typical structure, but the function may return, as we have seen, additional functions, such as plotting functions, or extra information, such as the size of the problem. As an example, le us implement a new problem, the _Maximum Clique Problem_. Briefly, this problem consists in, given a graph, finding the biggest subgraph that is complete. For the codification of the solution we will use a logical vector the size of the number of nodes in the graph and where a `TRUE` in the i-th element means tha the i-th node is included in the solution; a `FALSE` value will indicate that the node is not in the solution.

The instances of the problem will be defined by a graph. In our case, as in the rest of the graph problems in the package, we will use **igrph** to handle the graphs. As it can be interesting to explore some search strategies, we will also include a penalization factor in the evalution. This factor will be multiplied by the number of missing edges for the solution to be a clique and will be applied to the evaluation function.


```{r problem_example, echo=FALSE}
maximumCliqueProblem <- function(graph, penalization=0) {
  
  if (is.directed(graph)) {
      stop ("Only undirected graphs can be used for this problem")
  }
  graph <- simplify(graph)
  
  check <- function(solution) {
    if (class(solution)!="logical") {
      stop ("The solutions for the Maximum Clique Problem have to be logical vectors")
    }
    if (length(solution)!=vcount(graph)) {
      stop ("The solution has to have as many positions as nodes in the graph (", length(V(graph)), ")")
    }
  }
  
  evaluate <- function(solution) {
    check(solution)
    return(-1 * sum(solution))
  }
  
  valid <- function(solution) {
    check(solution)
    # Valid iff the induced subgraphs is complete
    sg <- induced.subgraph(graph, V(graph)[solution])
    num.nodes <- vcount(sg)
    num.edges <- ecount(sg)
    return(num.edges == num.nodes*(num.nodes-1)/2)
  }
  
  correct <- function(solution) {
    check(solution)
    while(!valid(solution)) {
      # Take the node in the current solution that has the least links and remove it from the solution
      sg <- induced.subgraph(graph, V(graph)[solution])
      degrees <- degree(sg, V(sg))
      id <- which.min(degrees)
      solution[which(solution)[id]] <- FALSE
      sum(solution)
    }
    return(solution)
  }
  
  plotSolution <- function(solution,  node.size=5, ...) {
    nodes <- V(graph)[solution]
    edgeInClique <- function (i) {
      return(all(get.edges(graph, E(graph)[i]) %in% nodes))
    }
    clique.edges <- sapply(1:ecount(graph), FUN=edgeInClique)
    
    width <- rep(1, ecount(graph))
    width[clique.edges] <- 3
    E(graph)$width <- width
    
    col.edge <- rep("lightgray", ecount(graph))
    col.edge[clique.edges] <- "black"
    E(graph)$color <- col.edge
    
    col.vertex <- rep("white", length(solution))
    col.vertex[solution] <- "black"
    V(graph)$color <- col.vertex
    plot.igraph(graph, vertex.size=node.size, 
                vertex.label=NA, edge.arrow.mode="-")
  }
  
  return(list(evaluate=evaluate, valid=valid, correct=correct, plotSolution=plotSolution))
}
```


As the codification may include non valid solutions, we will also include the `valid` and `correct` functions. Let us check the functions one by one. First of all, we need to set define the function inside which the required functions will be implemented, including any parameter we will need (a graph and a penalization factor in our example).

```{r problem_example, echo=1, eval=FALSE}
```

As  first step, we will do some checks in the graph and, to avoid problems, we will remove any loop or multiple edge.

```{r problem_example, echo=3:6, eval=FALSE}
```

Now, for convenience, we will define a function to check that a solution has the codification and size required to be a solution. Note that this function does not check tha the solution is valid (we will have another function for that), it only checks the validity of the codification. In the rest of the functions we will use `check` to verify the codification of a solution prior to be used.

```{r problem_example, echo=7:15, eval=FALSE}
```

As we have mentioned, for every single problem we need an evaluation function. In this case the function is quite simple, as it only counts the number of nodes in the solution (i.e., the clique). However, there is an important point to have in mind, **the algorithms in the package have been implemented to minimize the objective function**. As the goal in our problem is to maximize the size of the clique, we will **change the sign** of the evaluation.

```{r problem_example, echo=17:20, eval=FALSE}
```

Our problem has constraints, so we need to implement, at least, a `valid` function. This function has to check that the subgraph induced by the solution (i.e., that containing only the selected nodes) is a complete graph.

```{r problem_example, echo=21:28, eval=FALSE}
```

In this example we will include a simple greedy procedure to transform an invalid solution into a valid one. Briefly, the idea is iteratively removing from the subgraph the node with the less edges until the resulting subgraph is complete.

```{r problem_example, echo=30:42, eval=FALSE}
```

Additionally, we will include in our problem a handy function that will plot any solution. The graph will be plotted and the nodes and edges in the solution (the clique) will be highlighted in black.


```{r problem_example, echo=44:65, eval=FALSE}
```

Finally, we close the function returning a list with the functions we have just generated.

```{r problem_example, echo=66:67, eval=FALSE}
```

Once we have our function we can create new problems, check and correct solutions and plot them.

```{r use_example, eval=TRUE}
num.nodes <- 25
rnd.graph <- random.graph.game(n=num.nodes, p.or.m=0.5)
mcp <- maximumCliqueProblem(rnd.graph)
rnd.sol <- runif(num.nodes) > 0.5
mcp$valid(rnd.sol)
sol <- mcp$correct(rnd.sol)
mcp$valid(sol)
mcp$evaluate(sol)
mcp$plotSolution(sol)
```

More importantly, we use the functions we have just generated to apply any of the search algorithms included in the package (if you do not understand the code bellow, please see the vignettes devoted to the search algorithms). 

```{r use_example_2, eval=TRUE}

rnd.graph <- random.graph.game(n=num.nodes, p.or.m=0.75)
mcp <- maximumCliqueProblem(rnd.graph)
init.sol <- rep(FALSE, num.nodes)

args <- list()
args$evaluate         <- mcp$evaluate
args$initial.solution <- init.sol
args$neighborhood     <- flipNeighborhood(init.sol)
args$selector         <- greedySelector
args$non.valid        <- "correct"
args$valid            <- mcp$valid
args$correct          <- mcp$correct
args$resources        <- cResource(time=15)

res <- do.call(basicLocalSearch, args)

```


